Oak resilience to drought and land-use show how the ecological and geographical rear-edges do not necessarily meet today

### 2019-09-25 man/ecosystems/ms_ecosystems.Rmd
update

---
title: "Oak resilience to drought and land-use show how the ecological and geographical rear-edges do not necessarily meet today"
output:
  word_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 6
    highlight: null
    reference_docx: ../../templates/template_ms.docx
bibliography: ../../refs/references.bib
csl: ../../refs/ecosystems.csl
---

@@

## **Short Title**
Relict oak resilience at the rear-edge

## Authors [^\*^] 
A.J. Pérez-Luque^1,2,a^; G. Gea-Izquierdo^3,b^ and Zamora, R.^1,2,c^

^1^ Instituto Interuniversitario de Investigación del Sistema Tierra en Andalucía (CEAMA), Universidad de Granada. Avda. del Mediterráneo s/n, E-18006 Granada, Spain

^2^ Grupo de Ecología Terrestre, Departamento de Ecología, Facultad de Ciencias, Universidad de Granada, Avda. Fuentenueva s/n, E-18071 Granada, Spain

^3^ INIA-CIFOR. Ctra. La Coruña km 7.5. E-28040 Madrid, Spain

^a^ ajperez@ugr.es

^b^ gea.guillermo@inia.es

^c^ rzamora@ugr.es

@@

## Manuscript highlights
* *Quercus pyrenaica* rear-edge forests showed high resilience at tree and stand-levels
* The responses to climate and resilience followed a water-stress gradient
* Trees and stand expressed high sensitivity to drought and land-use legacies

@@

## Keywords
extreme drought, resilience, rear-edge, *Quercus pyrenaica*, tree-growth, dendrocronology, remote-sensing

# Introduction

Resumir párrafos de Vilá-Cabrera (los que ha puesto RZ)

Assessment of drought impacts on forest ecosystems has gained much attention in last decades [@Allen2010; @Clark2016], and the functional response of vegetation to drought has been analyzed at global and local scales [*e.g.*, @VicenteSerrano2013; see @MartinezVilalta2016 for a revision]. There has been an increase in the frequency and severity of drought events in the last decades, along with a trend towards drier summers particularly for southern Europe [@VicenteSerrano2014; @Stagge2017]. This is especially critical, since climate change projections for the Mediterranean area forecast that climate extreme events will become more frequent and severe [@Spinoni2017a]. In a global-change context, the impacts of drought on vegetation need to be evaluated considering the interactions with other drivers of change [@DoblasMiranda2017]. This is especially relevant for areas with a long history of landscape modification, such us the Mediterranean region, where land-use change plays a significant role in current ecosystem ecology [@Navarro2013], considering also that the interactions of these two factors, namely drought and land-use change, are crucial for Mediterranean forests [@DoblasMiranda2017].

The impacts of drought are also especially relevant for populations considered to be located in the rear-edge (either latitudinal or altitudinal along climatic gradients) of their distribution, where climatic conditions mark the species-limits and the populations are likely to be more sensitive to small variations on climate and the microtopography [@Hampe2005]. The response of rear-edge populations to variations on climate could be useful to estimate the adaptation of the species to predicted climatic changes [@Jump2010]. Yet it is crucial to accurately characterize the rear-edge of species. This can be particularly misleading in ecosystems with strong land-use legacies because it is established from current species distribution [@VilaCabreraJump2019GreaterGrowth]. Furthermore, it becomes even more important for relict populations driven by historical land-uses if they harbour high levels of intraspecific genetic diversity [@Valbuena2013].

Many studies have analyzed the effects of drought on tree species which southernmost limit of their distribution is encountered in the Mediterranean [*e.g.*, @SanchezSalguero2012; @Dorado2017AFM], and some focused on the resilience to drought of rear-edge populations [*e.g.*, @Herrero2014]. Resilience is the capacity of an ecosystem, community or individual to recover its pre-disturbance structure and function after a disturbance [@Holling1973]. The assessment of resilience to climate disturbances like droughts provides critical information about the capacity of the forests to maintain their structure and provide valuable ecosystem services.

Understanding the vegetation response to drought requires of integrative and multidisciplinary strategies to assess the impacts of disturbances on forest at several scales [@Hartman2018]. The resilience to drought of species is better assessed when several approaches like remote sensing and dendrocronology are combined simultaneously. Dendroecological estimates of growth (*i.e.* tree-ring width) are a widely used proxy to study tree vitality and the analysis of annual-tree ring widths can be used to study changes in growth as response to drought at the individiual tree-scale [@Fritts1976]. The combination of both approaches (remote sensing and dendroecology), has been used to assess the effects of droughts on vegetation along ecological gradients [*e.g.* @VicenteSerrano2013; @Coulthard2017], and also for the analysis of growth resilience to drought on several tree species [@Gazol2018]. Remote sensing can be used for studying droughts impacts on ecosystems, hence the stand level [*e.g.* @Zhang2013]. Vegetation indices obtained from satellite are used as a proxy for assessing vegetation functioning and aboveground net primary production [@AlcarazSegura2009], and changes in the patterns of these indices are considered as an indicator of the forest response to external environmental factors, such us drought [*e.g.* @VicenteSerrano2013]. Drought impacts on vegetation using remote sensing have been analyzed at different scales. For instance, a global-scale study showed a stronger influence of drought for drier biomes [@VicenteSerrano2013] whereas for Mediterranean area, studies at regional and local scales found a large spatial and seasonal variability of the effects of drought on vegetation [@Gouveia2017]. Several works assessed vegetation responses in term of resistance and resilience to drought events highlighting a stronger response of tree-ring (tree-level) to drought variability compared to the greenness (ecosystem-level) [*e.g.* @PenaGallardo2018; @Gazol2018]. Nonetheless, it is crucial to know if the responses at the tree level differ from responses at the ecosystem level and characterize the spatial variability of this response in rear-edge populations.

@@

# Materials and methods

## Tree species and study site

Pyrenean oak (*Quercus pyrenaica* Willd.) forests extend through south-western France and the Iberian Peninsula reaching its southern limit in mountains areas of northern Morocco [@Franco1990]. In the Iberian Peninsula these forests live on siliceous soils under meso-supramediterranean and mesotemperate areas and subhumid, humid and hyperhumid ombroclimate. *Q. pyrenaica* is a deciduous species that requires over 650 mm of annual precipitation and some summer precipitation. It is a submediterranean species, hence with lower drought tolerance than genuine evergreen Mediterranean taxa [@delRio2007].

The forests of this species reach their southernmost European limit in mountains from Southern Andalusia like Sierra Nevada (37°N, 3°W), a high-mountain range with elevations between 860 m and 3482 m *a.s.l.*. The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought (July-August) but with marked variability set by altitude. There are eight Pyrenean oak patches (2400 Has) identified in this mountain range (Figure 1), ranging between 1100 and 2000 m *a.s.l.* and often associated to major river valleys. Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species [@Olalde2002]. Today, *Q. pyrenaica* woodlands in this mountain region represent a rear edge of their habitat distribution [@Hampe2005]. They are the richest vascular-plant forest formation of Sierra Nevada, sheltering several endemic and endangered plant species [@Loriteetal2008ABG]. These relict forests undergone intensive anthropic use in history [@CamachoOlmedo2002; @Valbuena2017]. Furthermore, conservation status of this species for southern Spain is considered "Vulnerable" and it is expected to suffer the impact of climate change, reducing its suitable habitats in the rear-edge [@GeaIzquierdo2013].

## Drought episodes

The Iberian Peninsula suffered several extreme drought episodes in the last three decades (*e.g*. 1994-1995, 1999, 2005, 2012) [@VicenteSerrano2014]. The 2005 and 2012 drought events have been documented among the worst droughts in the last decades for the southern Iberian Peninsula [@Pascoa2017] and they were characterized as extreme drought in our climatic data (Figures S1-S2; Table S3). We focused on these two drought events because they were included in the period where there is availability of high-spatial resolution of remote sensing information (MODIS started on 2000; see below). Nevertheless, for radial growth-time series, a greater number of older drought events were also analyzed to contextualize results obtained in 2005 and 2012 drought events and analyze forest resilience to drought in a longer term (see Table S3). The identification of a drought event were done using the Standardized Precipitation-Evapotranspiration Index (SPEI) [@VicenteSerrano2010] following similar procedure as proposed by @Spinoni2015. We used SPEI-12 months scale for all 0.5º grid cells covering Sierra Nevada obtained from Global SPEI Database (http://spei.csic.es/database.html). A severe drought event starts in the month when SPEI falls below the threshold of -1.28 [@Spinoni2017a; @Pascoa2017]. A drought event is considered only when SPEI values are below that threshold for at least two consecutive months. Several indicators were computed for each drought event: *duration* as the number of consecutive months with the SPEI lower than a certain threshold; *severity* of a drought event is the sum of the SPEI values (absolute values) during the duration of the drought event; the *intensity* and the *Lowest SPEI* refer to the mean and lowest value of SPEI respectively during the drought event.

## Greenness data to assess ecosystem resilience

Vegetation greenness of *Q. pyrenaica* was characterized by means of the *Enhanced Vegetation Index* (EVI), derived from MOD13Q1 product of the MODIS (*Moderate Resolution Imaging Spectroradiometer*) sensor. MODIS EVI Data (Collection 6) were obtained for the period 2000 - 2016. EVI data consists of 16-day maximum value composite images (23 per year) of the EVI value with a spatial resolution of 250 m x 250 m. We selected the pixels covering the distribution of *Q. pyrenaica* forests in Sierra Nevada (*n* = 928 pixels). Values affected by high content of aerosols, clouds, snow and shadows, were filtered out following recommendations for mountain regions [@Reyes2015].

Annual mean EVI ($EVI_{mean}$) as a surrogate of mean annual primary production was computed for each pixel for the period 2000 - 2016. The EVI standardized anomaly ($EVI~sa$) was computed pixel-by-pixel, to minimize bias in the evaluation of anomalies and provide more information about the magnitude of anomalies [@Samanta2012]. For each pixel, an annual EVI value was obtained by averaging EVI valid values. Then, the standardized anomaly was computed as:

$$\mathrm{EVI_{sa,\mathit{i}}}= \frac{\mathrm{EVI_{mean,\mathit{i}}-EVI_{mean,ref}}}{\sigma_{\mathrm{ref}}}$$
where $\mathrm{EVI_{sa,\mathit{i}}}$ is the EVI standardized anomaly for year $i$;  $\mathrm{EVI_{mean,\mathit{i}}}$ the annual mean value of EVI for year $i$; $\mathrm{EVI_{mean,ref}}$ the average of the annual EVI values for the period of reference 2000-2016 (all except year $i$); and $\sigma_{\mathrm{ref}}$ the standard deviation for the reference period. Each pixel was categorized according the EVI standardized anomalies as "greening" ($EVI~sa > 1$), "browning" ($EVI~sa <- 1$) or "no changes" ($-1 > EVI~sa > 1$)[@Samanta2012].

Rather than other vegetation indices like the NDVI, $EVI_{mean}$ was chosen since it is highly stable under the use of any filter [@Reyes2015] and because it showed high significant correlations with annual ($r$ = 0.81) and seasonal EVI values ($r_{spring}$ = 0.76 and $r_{summer}$ = 0.88). For exploratory purposes, temporal trends of $EVI_{mean}$ were explored at a pixel scale, using Mann–Kendall nonparametric test.

## Field sampling and dendrochronological methods to assess tree individual resilience

Tree sampling was carried out during autumn 2016. Trees were sampled at two locations in contrasting N-S slopes of Sierra Nevada: San Juan (SJ), a xeric site located at the northern aspect; and Cáñar (CA), a more humid site located at the southern aspect (Figure 1; Table 1). For the southern site two elevations were sampled: CA-Low and CA-High, which constitute the current altitudinal rear-edge (CA-Low) and tree-line (CA-High), respectively in the sampled aspect Sampling sites followed a moisture gradient: SJ < CA-Low < Ca-High (Table 1). All the sites were oak monospecific and representative of the population clusters identified for the species in this mountain range [@PerezLuque2015]. In each site between 15 and 20 trees from either the single dominant-codominant layer (CA) or the open canopy (SJ) were randomly selected. Two cores of 5 mm of diameter were taken from each tree at breast heigth (1.3 m) using an increment borer. Diameter at breast height (DBH) and total height were measured for each tree. In addition, stand competition affecting target trees was assessed by recording distance, azimuth, DBH, species and total height of all neighboring living trees with DBH > 7.5 cm within a circular plot of 10 m radius. Several competition indices were calculated: the distance independent indices *density* ($\mathrm{trees \cdot ha^{-1}}$), and *basal area* (BA, $\mathrm{m^{2} \cdot ha^{-1}}$); and the distance dependent index size ratio proportional to distance (*srd*) as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$ [see @GeaIzquierdo2009 for more details]. Differences between sites for height, DBH and competition indices were analysed using non-parametric Kruskal-Wallis rank sum tests. When significant differences were observed, we run multiple comparisons using the Dunn's-test with Bonferroni adjustment to correct for significance.

@@

### Climate and growth

Climate data were obtained from the European Daily High-Resolution Observational Gridded Dataset (E-OBS v16) [@Haylock2008]. Monthly precipitation and minimum and maximum temperatures were obtained at a 0.25 x 0.25 º resolution for the 1950-2016 period. We selected grid cells covering each sampled sites. The Standardized Precipitation-Evapotranspiration Index (SPEI) with a temporal scale of 6 months was used to characterize the drought conditions for the period 1961-2014. Relationships between RWI and BAI site chronologies and climatic variables were assessed using bootstrapped Pearson's correlations estimated using `treeclim` [@Zang2015].

### Disturbance analyses 
Disturbance chronologies were built using tree-ring width to identify abrupt and sustained increases (release events from competition) or decreases (supressions) in radial growth [@Nowacki1997] as indirect estimates of possible disturbance events (e.g. logging, drought induced neighbor mortality) in the past. Growth changes (GC) were calculated for the individual tree-ring series using a 10-year running window as either positive (PGC) or negative (NGC) growth changes:

$$\mathrm{\% GC} = \left [ \frac{\left ( M1 - M2 \right )}{M2}\right] \times 100 $$
where $M1$ is the preceding 10-year median and $M2$ is the subsequent 10-year median. Medians were used since they are a more robust estimator of central tendency than means [@Rubino2004]. Site disturbance chronologies were constructed by averaging the individual disturbances series annually. To separate growth peaks produced by disturbance events and expressing stand-wise disturbances from those by climate, we considered a threshold of 50 % of GC and more than 50% of the individual trees displaying the same growth changes [*e.g.* @GeaIzquierdo2014]. In addition, the forest and management history of our sampling sites was inferred from a detailed analysis of historical land-use changes. For this purpose, an exhaustive review of existing abundant historical documents was done to compile information about socio-economical activities affecting the studied forests (Table S4).

@@

## Assessing ecosystem and tree individual resilience to drought

To evaluate the effects of drought events on ecosystem resilience (using greenness data) and tree individual resilience (using BAI data) we used resilience indices proposed by @Lloret2011: resilience (*Rs*), resistance (*Rt*), and recovery (*Rc*). The Resistance index, estimated as the ratio between performance during and before the disturbance ($Rt=Drought / PreDrought$), quantifies the severity of the impact of the disturbance in the year it occurred. The Recovery index, computes as the ratio between performance after and during disturbance ($Rc = PostDrought / Drought$), represents the ability to recover from disturbance relative to its severity. Finally, the Resilience index ($Rs = PostDrought / PreDrought$) is the capacity to reach pre-disturbance performance levels. The values of these indices were computed for tree growth (BAI) and greenness (EVI mean) during each drought event. The predrought and postdrought values of each target variable (*i.e.* BAI or EVI) were computed as the mean value during a period of three years before and after the drought event respectively. A period of three years was chosen because we found similar results comparing periods of two, three and four years (Figure S6) and this length was used in other studies [*e.g.* @Gazol2018]. To put in context in a longer-term the resilience observed for 2005 and 2012 drought events, as well as to test whether the resilience indices were ecologically meaningful, resilience metrics for BAI data were additionally computed for the most severe drought events since 1950 and compared with drought severity (Figure 6, Table S3).

## Statistical analysis

For each of the three resilience indices studied we used robust two-way ANOVAs to test for differences between drought events (2005 and 2012) and the studied oak populations (northern and southern aspects). These tests were used because original and log-transformed data did not follow the assumptions of normality and homogeneity of variance [@Wilcox2012]. Robust measures of central tendency (M-estimator based on Huber's Psi) were used since they were close to the mean value in all cases [@Wilcox2012]. When running the robust ANOVA test, data were bootstrapped 3000 times and trimmed automatically to control the potential influence of outliers. Post-hoc differences were assessed pairwise using a similar bootstrap test. All the robust ANOVA and post-hoc tests were carried out using the `WRS2` package. The level of significance was set to 0.05 and adjusted for multiple comparisons.

# Results

## Time trends in vegetation greenness

The analysis of greenness time trends showed that EVI in 78.9 % of the pixels experienced a positive trend for the 2000-2016 period. The lowest values of EVI standardized anomalies for the studied period were recorded during the 2005 drought, and the minimum EVI values were expressed in the northern (dry) population (Figure 2). A "browning" episode ($\mathrm{EVI_{sa}} < -1$) was observed during this drought event, whereas no changes in greenness were observed in response to the 2012 drought (Figure 3).

## Analysis of radial growth trends and disturbances

>>>>

---
title: "Oak resilience to drought and land-use show how the ecological and geographical rear-edges do not necessarily meet today"
output:
  word_document: 
    fig_caption: yes
    fig_height: 6
    highlight: null
    reference_docx: ../../templates/template_ms.docx
bibliography: ../../refs/references.bib
csl: ../../refs/ecosystems.csl
---

@@

### Short title: *Relict oak resilience at the rear-edge*

### Authors [^\*^] 
A.J. Pérez-Luque^1,2,a^; G. Gea-Izquierdo^3,b^ and Zamora, R.^1,2,c^

^1^Instituto Interuniversitario de Investigación del Sistema Tierra en Andalucía (CEAMA), Universidad de Granada. Avda. del Mediterráneo s/n, E-18006 Granada, Spain. ^2^Grupo de Ecología Terrestre, Departamento de Ecología, Facultad de Ciencias, Universidad de Granada, Avda. Fuentenueva s/n, E-18071 Granada, Spain. ^3^INIA-CIFOR. Ctra. La Coruña km 7.5. E-28040 Madrid, Spain

^a^ ajperez@ugr.es ^b^ gea.guillermo@inia.es ^c^ rzamora@ugr.es

@@

### Manuscript highlights
* *Quercus pyrenaica* rear-edge forests showed high resilience at tree and stand-levels
* The responses to climate and resilience followed a water-stress gradient
* Trees and stand expressed high sensitivity to drought and land-use legacies

@@

### Keywords 
extreme drought, resilience, rear-edge, *Quercus pyrenaica*, tree-growth, dendrocronology, remote-sensing

## Introduction
The response of species to changing environments is likely to be determined largely by population responses at range margins [@Hampe2005]. Peripheral populations are usually considered more vulnerable compared with populations occurring at the center of a species' range (*i.e.* centre-periphery hypothesis) [@Sagarin2002; @Pironon2016]. It has been assumed that rear-edge populations have a concordance between geographical and ecological marginality, with lower performance and thus higher risk of extinction than those populations at the core of the species' range [@VilaCabrera2019; @Pironon2016]. Nonetheless, recent reviews found species- and population- specific responses highlight that this hypothesis not always fulfill [@Sexton2009; @Abeli2014]. In this respect, to fully understand changes in distribution and abundance of species as a consequence of global change we need to know in which environmental conditions geographical marginallity does not correspond with ecological marginallity [@VilaCabreraJump2019GreaterGrowth].

Climate change is expected to cause important changes in the distribution and abundance of plant communities, and there are already obvious signs of processes associated with increased intensity and duration of droughts [@Allen2010]. In this respect, several studies have warned on the rapid extinction of plant species in the near future, especially in mountain areas of southern Europe [*e.g.* @Engler2011]. Under this climatic change context, it is reasonable to expect that population loss and range retractions should be seen in the most low-latitude, drought-prone areas of a species' distribution, *i.e.* the rear edge [@Hampe2005].

Assessment of drought impacts on forest ecosystems has gained much attention in last decades [@Allen2010], and the functional response of vegetation to drought has been analyzed at global and local scales [*e.g.*, @VicenteSerrano2013; @MartinezVilalta2016]. There has been an increase in the frequency and severity of drought events in the last decades, along with a trend towards drier summers particularly for southern Europe [@VicenteSerrano2014; @Stagge2017]. This is especially critical, since climate change projections for the Mediterranean area forecast that climate extreme events will become more frequent and severe [@Spinoni2017a]. In a global-change context, the impacts of drought on vegetation need to be evaluated considering the interactions with other drivers of change [@DoblasMiranda2017]. This is especially relevant for areas with a long history of landscape modification, such us the Mediterranean region, where land-use change plays a significant role in current ecosystem ecology [@Navarro2013], considering also that the interactions of drought and land-use change are crucial for Mediterranean forests [@DoblasMiranda2017].

The impacts of drought are also especially relevant for populations considered to be located in the rear-edge (either latitudinal or altitudinal along climatic gradients) of their distribution, where climatic conditions mark the species-limits and the populations are likely to be more sensitive to small variations on climate and the microtopography [@Hampe2005]. The response of rear-edge populations to variations on climate could be useful to estimate the adaptation of the species to predicted climatic changes [@Jump2010]. Yet it is crucial to accurately characterize the rear-edge of species. This can be particularly misleading in ecosystems with strong land-use legacies because it is established from current species distribution [@VilaCabrera2019]. Furthermore, it becomes even more important for relict populations driven by historical land-uses if they harbour high levels of intraspecific genetic diversity [@Valbuena2013].

Many studies have analyzed the effects of drought on tree species which southernmost limit of their distribution is encountered in the Mediterranean [@SanchezSalguero2012; @Dorado2017AFM], and some focused on the resilience to drought of rear-edge populations [*e.g.*, @Herrero2014]. Resilience is the capacity of an ecosystem, community or individual to recover its pre-disturbance structure and function after a disturbance [@Holling1973]. The assessment of resilience to climate disturbances like droughts provides critical information about the capacity of the forests to maintain their structure and provide valuable ecosystem services.

Understanding the vegetation response to drought requires of integrative and multidisciplinary strategies to assess the impacts of disturbances on forest at several scales [@Hartman2018]. The resilience to drought of species is better assessed when several approaches like dendrocronology and remote sensing are combined simultaneously. Dendroecological estimates of growth (*i.e.* tree-ring width) are a widely used proxy to study tree vitality and the analysis of annual tree-ring widths can be used to study changes in growth as response to drought at the individiual tree-scale [@Fritts1976]. The combination of both approaches (remote sensing and dendroecology), has been used to assess the effects of droughts on vegetation along ecological gradients [*e.g.* @VicenteSerrano2013; @Coulthard2017], and also for the analysis of growth resilience to drought on several tree species [@Gazol2018]. Remote sensing can be used for studying droughts impacts on ecosystems, hence the stand level [*e.g.* @Zhang2013]. Vegetation indices obtained from satellite are used as a proxy for assessing vegetation functioning and aboveground net primary production [@AlcarazSegura2009], and changes in the patterns of these indices are considered as an indicator of the forest response to external environmental factors, such us drought [*e.g.* @VicenteSerrano2013]. Drought impacts on vegetation using remote sensing have been analyzed at different scales [@VicenteSerrano2013]. For instance, a large spatial and seasonal variability of the effects of drought on vegetation has been documented for Mediterranean area [@Gouveia2017]. Several works assessed vegetation responses in term of resistance and resilience to drought events highlighting a stronger response of tree-ring (tree-level) to drought variability compared to the greenness (ecosystem-level) [*e.g.* @PenaGallardo2018; @Gazol2018]. Nonetheless, it is crucial to know if the responses at the tree level differ from responses at the ecosystem level and characterize the spatial variability of this response in rear-edge populations.

@@

## Materials and methods 
### Tree species and study site 
Pyrenean oak (*Quercus pyrenaica* Willd.) forests extend through south-western France and the Iberian Peninsula reaching its southern limit in mountains areas of northern Morocco [@Franco1990]. In the Iberian Peninsula these forests live on siliceous soils under meso-supramediterranean and mesotemperate areas and subhumid, humid and hyperhumid ombroclimate. *Q. pyrenaica* is a deciduous species that requires over 650 mm of annual precipitation and some summer precipitation. It is a submediterranean species, hence with lower drought tolerance than genuine evergreen Mediterranean taxa [@delRio2007].

The forests of this species reach their southernmost European limit in mountains from Southern Andalusia like Sierra Nevada (37°N, 3°W), a high-mountain range with elevations between 860 and 3482 m *a.s.l.*. The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought but with marked variability set by altitude. Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species [@Olalde2002]. There are eight Pyrenean oak patches (2400 Has) identified in this mountain range (Figure 1), ranging between 1100 and 2000 m *a.s.l.* and often associated to major river valleys. Today, *Q. pyrenaica* woodlands in this mountain region represent a rear edge of their habitat distribution [@Hampe2005]. They are the richest vascular-plant forest formation of Sierra Nevada, sheltering several endemic and endangered plant species [@Loriteetal2008ABG]. These relict forests undergone intensive anthropic use in history [@CamachoOlmedo2002]. Furthermore, conservation status of this species for southern Spain is considered "Vulnerable" and it is expected to suffer the impact of climate change, reducing its suitable habitats in the rear-edge [@GeaIzquierdo2013].

### Drought episodes
The Iberian Peninsula suffered several extreme drought episodes in the last three decades (*e.g*. 1994-1995, 1999, 2005, 2012) [@VicenteSerrano2014]. The 2005 and 2012 drought events have been documented among the worst droughts in the last decades for the southern Iberian Peninsula [@Pascoa2017] and they were characterized as extreme drought in our climatic data (Figure S1; Table S3). We focused on these two drought events because they were included in the period where there is availability of high-spatial resolution of remote sensing information (MODIS started on 2000; see below). Nevertheless, for radial growth-time series, a greater number of older drought events were also analyzed to contextualize results obtained in 2005 and 2012 drought events and analyze forest resilience to drought in a longer term (see Table S3). The identification of a drought event were done using the Standardized Precipitation-Evapotranspiration Index (SPEI) [@VicenteSerrano2010] following similar procedure as proposed by @Spinoni2015. We used SPEI-12 months scale for all 0.5º grid cells covering Sierra Nevada obtained from Global SPEI Database (http://spei.csic.es/database.html). A severe drought event starts when SPEI falls below the threshold of -1.28 [@Spinoni2017a; @Pascoa2017]. A drought event is considered only when SPEI values are below that threshold for at least two consecutive months. For each drought event we computed: the *duration*, *i.e.* number of consecutive months with the SPEI lower than a certain threshold; the *severity* as the sum of the SPEI values (absolute values) during the duration of the drought event; the *intensity* and the *Lowest SPEI* refer to the mean and lowest value of SPEI respectively during the drought event.

### Greenness data to assess ecosystem resilience
Vegetation greenness of *Q. pyrenaica* was characterized by means of the *Enhanced Vegetation Index* (EVI), derived from MOD13Q1 product of the MODIS (*Moderate Resolution Imaging Spectroradiometer*) sensor. EVI data consists of 16-day maximum value composite images (23 per year) of the EVI value with a spatial resolution of 250 m x 250 m. MODIS EVI Data (Collection 6) were obtained for the period 2000 - 2016. We selected the pixels covering the distribution of *Q. pyrenaica* forests in Sierra Nevada (*n* = 928 pixels). Values affected by high content of aerosols, clouds, snow and shadows, were filtered out following recommendations for mountain regions [@Reyes2015].

Annual mean EVI ($EVI_{mean}$) as a surrogate of mean annual primary production was computed for each pixel for the period 2000 - 2016. The EVI standardized anomaly ($EVI_{sa}$) was computed pixel-by-pixel, to minimize bias in the evaluation of anomalies and provide more information about the magnitude of anomalies [@Samanta2012]. For each pixel, an annual EVI value was obtained by averaging EVI valid values. Then, the standardized anomaly was computed as:

$$\mathrm{EVI_{sa,\mathit{i}}}= \frac{\mathrm{EVI_{mean,\mathit{i}}-EVI_{mean,ref}}}{\sigma_{\mathrm{ref}}}$$
where $\mathrm{EVI_{sa,\mathit{i}}}$ is the EVI standardized anomaly for year $i$;  $\mathrm{EVI_{mean,\mathit{i}}}$ the annual mean value of EVI for year $i$; $\mathrm{EVI_{mean,ref}}$ the average of the annual EVI values for the period of reference 2000-2016 (all except year $i$); and $\sigma_{\mathrm{ref}}$ the standard deviation for the reference period. Each pixel was categorized according the EVI standardized anomalies as "greening" ($EVI_{sa} > 1$), "browning" ($EVI_{sa} <- 1$) or "no-changes" ($-1 > EVI_{sa} > 1$)[@Samanta2012].

Rather than other vegetation indices like the NDVI, $EVI_{mean}$ was chosen since it is highly stable under the use of any filter [@Reyes2015] and because it showed high significant correlations with annual ($r$= 0.81) and seasonal EVI values ($r_{spring}$= 0.76 and $r_{summer}$= 0.88). For exploratory purposes, temporal trends of $EVI_{mean}$ were explored at a pixel scale, using Mann–Kendall nonparametric test.

### Field sampling and dendrochronological methods to assess tree individual resilience
Tree sampling was carried out during autumn 2016. Trees were sampled at two locations in contrasting N-S slopes of Sierra Nevada: San Juan (SJ), a xeric site located at the northern aspect; and Cáñar (CA), a more humid site located at the southern aspect (Figure 1; Table 1). For the southern site two elevations were sampled: CA-Low and CA-High, which constitute the current altitudinal rear-edge (CA-Low) and tree-line (CA-High), respectively in the sampled aspect. Sampling sites followed a moisture gradient: SJ < CA-Low < Ca-High (Table 1). All the sites were oak monospecific and representative of the population clusters identified for the species in this mountain range [@PerezLuque2015]. In each site between 15 and 20 trees from either the single dominant-codominant layer (CA) or the open canopy (SJ) were randomly selected. Two cores of 5 mm of diameter were taken from each tree at breast heigth (1.3 m) using an increment borer. Diameter at breast height (DBH) and total height were measured for each tree. In addition, stand competition affecting target trees was assessed by recording distance, azimuth, DBH, species and total height of all neighboring living trees with DBH > 7.5 cm within a circular plot of 10 m radius. Several competition indices were calculated: the distance independent indices *density* ($\mathrm{trees \cdot ha^{-1}}$), and *basal area* (BA, $\mathrm{m^{2} \cdot ha^{-1}}$); and the distance dependent index size ratio proportional to distance as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$ [see @GeaIzquierdo2009 for more details]. Differences between sites for height, DBH and competition indices were analysed using non-parametric Kruskal-Wallis rank sum tests. When significant differences were observed, we run multiple comparisons using the Dunn's-test with Bonferroni adjustment to correct for significance.

@@

### Climate and growth 
Climate data were obtained from the European Daily High-Resolution Observational Gridded Dataset (E-OBS v16) [@Haylock2008]. Monthly precipitation and minimum and maximum temperatures were obtained at a 0.25 x 0.25 º resolution for the 1950-2016 period. We selected grid cells covering each sampled sites. SPEI index with a temporal scale of 6 months was used to characterize the drought conditions for the period 1961-2014. Relationships between RWI and BAI site chronologies and climatic variables were assessed using bootstrapped Pearson's correlations estimated using `treeclim` [@Zang2015].

### Disturbance analyses 
Disturbance chronologies were built using tree-ring width to identify abrupt and sustained increases (release events from competition) or decreases (supressions) in radial growth [@Nowacki1997] as indirect estimates of possible disturbance events (*e.g.* logging, drought induced neighbor mortality) in the past. Growth changes (GC) were calculated for the individual tree-ring series using a 10-year running window as either positive (PGC) or negative (NGC) growth changes: $\mathrm{\% GC} = \left [ \frac{\left ( M1 - M2 \right )}{M2}\right] \times 100$, where $M1$ is the preceding 10-year median and $M2$ is the subsequent 10-year median.
```{r, eval=FALSE, echo=FALSE}
# Medians were used since they are a more robust estimator of central tendency than means [@Rubino2004].
```
Site disturbance chronologies were constructed by averaging the individual disturbances series annually. To separate growth peaks produced by disturbance events and expressing stand-wise disturbances from those by climate, we considered a threshold of 50 % of GC and more than 50% of the individual trees displaying the same growth changes [*e.g.* @GeaIzquierdo2014]. In addition, the forest and management history of our sampling sites was inferred from a detailed analysis of historical land-use changes. For this purpose, an exhaustive review of existing abundant historical documents was done to compile information about socio-economical activities affecting the studied forests (Table S4).

@@

### Assessing ecosystem and tree individual resilience to drought
To evaluate the effects of drought events on ecosystem resilience (using greenness data) and tree individual resilience (using BAI data) we used resilience indices proposed by @Lloret2011. The Resistance index estimated as the ratio between performance during and before the disturbance ($Rt=Drought / PreDrought$), quantifies the severity of the impact of the disturbance in the year it occurred. The Recovery index, computes as the ratio between performance after and during disturbance ($Rc = PostDrought / Drought$), represents the ability to recover from disturbance relative to its severity. Finally, the Resilience index ($Rs = PostDrought / PreDrought$) is the capacity to reach pre-disturbance performance levels. The values of these indices were computed for tree growth (BAI) and greenness (EVI mean) during each drought event. The predrought and postdrought values of each target variable (*i.e.* BAI or EVI) were computed as the mean value during a period of three years before and after the drought event respectively. A period of three years was chosen because we found similar results comparing periods of two, three and four years (Figure S4b) and this length was used in other studies [*e.g.* @Gazol2018]. To put in context in a longer-term the resilience observed for 2005 and 2012 drought events, as well as to test whether the resilience indices were ecologically meaningful, resilience metrics for BAI data were additionally computed for the most severe drought events since 1950 and compared with drought severity (Figure 5, Table S3).

### Statistical analysis 
For each of the three resilience indices studied we used robust two-way ANOVAs to test for differences between drought events (2005 and 2012) and the studied oak populations (northern and southern aspects). These tests were used because original and log-transformed data did not follow the assumptions of normality and homogeneity of variance [@Wilcox2012]. Robust measures of central tendency (M-estimator based on Huber's Psi) were used since they were close to the mean value in all cases [@Wilcox2012]. When running the robust ANOVA test, data were bootstrapped 3000 times and trimmed automatically to control the potential influence of outliers. Post-hoc differences were assessed pairwise using a similar bootstrap test. All the robust ANOVA and post-hoc tests were carried out using the `WRS2` package. The level of significance was set to 0.05 and adjusted for multiple comparisons.

## Results
### Time trends in vegetation greenness
The analysis of greenness time trends showed that EVI in 78.9 % of the pixels experienced a positive trend for the 2000-2016 period. The lowest values of EVI standardized anomalies for the studied period were recorded during the 2005 drought, and the minimum EVI values were expressed in the northern (dry) population (Figure 2a). A "browning" episode ($\mathrm{EVI_{sa}} < -1$) was observed during this drought event, whereas no changes in greenness were observed in response to the 2012 drought (Figure 2b).

### Analysis of radial growth trends and disturbances 
The trees of the southern population were older than those from the northern one. In addition, trees from the southern population at high altitude were taller and their growth was significantly faster than that of trees from the other two sites (Table 1, Figure 3) and despite competition measured as plot basal area was greatest in CA-High (Table 1). Growth and height of trees from the northern and the low-elevation southern population were similar (Figures 3 and S4a). Only trees from the southern sites showed significant positive growth trends since the late 1970s (Figure 3), being much more pronounced for the high-elevation, hence more humid, site (CA-High).
